### 2024-11-29 parts/appendix.tex
big corrections

\begin{proofof}[undecidable-non-negative:lem]
    Recall that by the undecidability of the Hilbert's Tenth Problem \cite[Problem 10 page 22]{HILB1902}, it is
    undecidable whether a given polynomial $P \in \Rel[\vec{X}]$ has a solution in
    $\Rel^n$ \cite{MATI1970,DAVIS1973}.
    The result follows by observing that
    checking whether a polynomial $P \in \Rel[\vec{X}]$ has no solutions in $\Rel^n$ if and only 
    if $P^2 - 1$ is \kl{non-negative}.

    Let us prove that checking whether $P \in \CoveredPoly$ is undecidable. Let
    $P \in \Rel[\vec{X}]$ be a polynomial, and let us write $P = P_+ - P_-$
    where both $P_+$ and $P_-$ belong to $\Nat[\vec{X}]$. Let $Z$ be a fresh
    indeterminate. We define $Q \defined P_+ (Z - 1) P_-$. By construction, $Q$
    has \kl{non-negative} \kl{maximal monomials}. Furthermore, $Q$ itself is
    \kl{non-negative} if and only if $\restr{Q}{Z = 0} = P$ is \kl{non-negative},
    since for every $n > 0$, $\restr{Q}{Z = n} \in \Nat[\vec{X}]$ is
    \kl{non-negative}. We conclude that 
    $P$ is \kl{non-negative} if and only if $Q$ is \kl{non-negative},
    if and only if $Q \in \CoveredPoly$.
\end{proofof}

\begin{proofof}[n-poly-combinatorics:lem]
	Let us fix a finite monoid $M$, a morphism $\mu \colon \Sigma^* \to M$, and
	let $\omega$ be an idempotent power for the finite monoid $M$,
	i.e., a number such that for all $x \in M$,
	$(x^{\omega})^2 = x^{\omega}$.
	We will prove by induction on $k \in \Nat_{\geq 1}$
	that for all
	$\pi \colon M^k \to \Nat$,
	the lemma holds with $s = 0$ for the function
	$f = \pi^\dagger$, i.e., such that
	$f(w)$ is the sum over all factorizations of $w$
	into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
	of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
	Since every \kl{$\Nat$-polyregular function}
	is obtained via some choice of $M, \mu, \pi$ (\cref{nat-rel-poly:def})
	we derive the desired result.

	In all cases, we will use a \kl{polynomial function} $q \colon \Nat^p \to
		\Sigma^*$ defined by $q(X_1, \dots, X_p) \defined \alpha_0 \prod_{i = 1}^p
		(u_i^{\omega \times X_i} \alpha_i)$ using some words $\alpha_i$ (for $0 \leq i
		\leq p$) and $u_i$ (for $1 \leq i \leq p)$ in $\Sigma^*$. That is, we will
	implicitly multiply all variables by $\omega$ in order to make the
	equations more readable. Note that when using an induction hypothesis, we
	will therefore have to check that the polynomial $q$ will have this
	specific form.

@@

	Now, given $(\ell, b, r)$ respecting the inequalities
	mentioned above,
	one can define the \kl{polynomial functions}
	$\CutPref{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
	and
	$\CutSuff{\ell,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
	so that
	$q(X_1, \dots, X_p) = \CutPref{\ell,r,b}(X_1, \dots, X_{\ell-1})
		q_{\ell, r}(X_{\ell} - b, \dots, X_p)$
	under the assumption that $X_\ell \geq b$:
	\begin{align*}
        \CutPref{\ell,r,b}(X_1, \dots, X_{\ell - 1})
        &\defined
        \alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1}
        u_\ell^{\omega b} (u_{\ell}^\omega)_{[1:r]}
        \quad .
        \\
		\CutSuff{\ell,r}(X_{1}, \dots, X_{p - \ell + 1})
        &\defined
        (u_\ell^\omega)_{[-r:]} u_\ell^{\omega X_1}
		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
		\quad .
	\end{align*}
	Remark that $\CutSuff{\ell,r}$ has all of its indeterminates
	multiplied by $\omega$, so that the induction hypothesis
	can be used.

@@

\begin{proofof}[polyrec-integer-strong:lem]
    Because $n P \in \CorrectPoly$,
    it is represented by a \kl{star-free $\Nat$-polyregular function} $f$ thanks to
    \cref{corrected-version:thm}. Let us now use \cref{n-sf-combinatorics:lem}
    to obtain a number $s \geq 1$ (that only depends on $P$) such
    that for all \kl{polynomial function} $q \colon \Nat^k \to \Sigma^*$, $g
    \circ q(X_1 + s, \dots, X_k + s)$ is a \kl{natural binomial function} $F$
    over $\Nat^k$. Let us write the
    \kl{natural binomial function} $F$ as follows $F(X_1, \dots, X_k) =
    \sum_{\card{\alpha} \leq d_2} e_{\alpha} \prod_{i = 1}^k \binom{X -
    p_{\alpha,i}}{\alpha(i)}$, with $e_{\alpha} \in \Nat$ for all
    $\card{\alpha} \leq d_2$.
    Now, because of the equality 
    $\binom{X - p - 1}{k} + \binom{X - p - 1}{k - 1} = \binom{X - p}{k}$,
    we conclude that we can find a $K \in \Nat$ and coefficients
    $f_{\alpha} \in \Nat$ such that 
    $F(X_1, \dots, X_k) =
    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \binom{X_i - K}{\alpha(i)}$.
    Now, remark that $P$ is an \kl{integer-valued} polynomial of
    degree $d$, and therefore one can write $P = \sum_{\card{\alpha} \leq d}
    c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$ with $c_{\alpha}
    \in \Rel$ for all $\card{\alpha} \leq d$
    (\cref{integer-binomial-polynomial:lem}). 
    As a consequence,
    $\lambda P = \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K - s}{\alpha(i)}$,
    and 
    by identifying coefficients in the previous two expressions, we obtain
    $c_{\alpha} = \lambda f_{\alpha}$ for all $\card{\alpha} \leq d$.
    Finally,
    $\lambda P(x_1 + s, \dots, x_k + s) = F(x_1, \dots, x_k)$ for all $\vec{x} \in \Nat^k$,
    and therefore
    $\lambda P (X_1 + s, \dots, X_k + s) =
    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$.
    In particular,
    we get
    \begin{equation*}
        \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)} =
        \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
    \end{equation*}
    And we conclude by term identification
    by term identification
    that $d = d_2$, and for all $\card{\alpha} \leq d$, $\lambda c_{\alpha} = f_{\alpha}$.
    As a consequence, $c_{\alpha} \in \Nat$ for all $\card{\alpha} \leq d$,
    and $P$
    is a \kl{natural binomial polynomial}.
    Remark that this reasoning remains true whenever we partially apply $P$ to some values
    in $\Nat$, as it amounts to 
    partially applying the function $f$. We have proven that 
    $P$ is a \kl{strongly natural binomial polynomial}.
\end{proofof}

>>>>

\begin{proofof}[undecidable-non-negative:lem]
    Recall that by the undecidability of the Hilbert's Tenth Problem \cite[Problem 10 page 22]{HILB1902}, it is
    undecidable whether a given polynomial $P \in \Rel[X_1, \dots, X_k]$ has a solution in
    $\Rel^k$ \cite{MATI1970,DAVIS1973}.
    The result follows by observing that
    checking whether a polynomial $P \in \Rel[\vec{X}]$ has no solutions in $\Rel^k$ if and only 
    if $P^2 - 1$ is \kl{non-negative}.

    Let us prove that checking whether $P \in \CoveredPoly$ is undecidable. Let
    $P \in \Rel[\vec{X}]$ be a polynomial, and let us write $P = P_+ - P_-$
    where both $P_+$ and $P_-$ belong to $\Nat[\vec{X}]$. Let $Z$ be a fresh
    indeterminate. We define $Q \defined P_+ + (Z - 1) P_-$. By construction, $Q$
    has \kl{non-negative} \kl{maximal monomials}. Furthermore, $Q$ itself is
    \kl{non-negative} if and only if $\restr{Q}{Z = 0} = P$ is \kl{non-negative},
    since for every $n > 0$, $\restr{Q}{Z = n} \in \Nat[\vec{X}]$ is
    \kl{non-negative}. We conclude that 
    $P$ is \kl{non-negative} if and only if $Q$ is \kl{non-negative},
    if and only if $Q \in \CoveredPoly$.
\end{proofof}

\begin{proofof}[n-poly-combinatorics:lem]
	Let us fix a finite monoid $M$, a morphism $\mu \colon \Sigma^* \to M$, and
	let $\omega$ be an idempotent power for the finite monoid $M$,
	i.e., a number such that for all $x \in M$,
	$(x^{\omega})^2 = x^{\omega}$.
	We will prove by induction on $k \in \Nat_{\geq 1}$
	that for all
	$\pi \colon M^k \to \Nat$,
	the lemma holds for the function
	$f = \pi^\dagger$, i.e., such that
	$f(w)$ is the sum over all factorizations of $w$
	into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
	of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
	Since every \kl{$\Nat$-polyregular function}
	is obtained via some choice of $M, \mu, \pi$ (\cref{nat-rel-poly:def})
	we derive the desired result.

	In all cases, we will use a \kl{pumping pattern} $q \colon \Nat^p \to
		\Sigma^*$ defined by $q(X_1, \dots, X_p) \defined \alpha_0 \prod_{i = 1}^p
		(u_i^{\omega \times X_i} \alpha_i)$ using some words $\alpha_i$ (for $0 \leq i
		\leq p$) and $u_i$ (for $1 \leq i \leq p)$ in $\Sigma^*$. That is, we will
	implicitly multiply all variables by $\omega$ in order to make the
	equations more readable. Note that when using an induction hypothesis, we
	will therefore have to check that the polynomial $q$ will have this
	specific form.

@@

	Now, given $(\ell, b, r)$ respecting the inequalities
	mentioned above,
	one can define the \kl{pumping patterns}
	$\CutPref{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
	and
	$\CutSuff{\ell,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
	so that
	$q(X_1, \dots, X_p) = \CutPref{\ell,r,b}(X_1, \dots, X_{\ell-1})
		q_{\ell, r}(X_{\ell} - b, \dots, X_p)$
	under the assumption that $X_\ell \geq b$:
	\begin{align*}
        \CutPref{\ell,r,b}(X_1, \dots, X_{\ell - 1})
        &\defined
        \alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1}
        u_\ell^{\omega b} (u_{\ell}^\omega)_{[1:r]}
        \quad .
        \\
		\CutSuff{\ell,r}(X_{1}, \dots, X_{p - \ell + 1})
        &\defined
        (u_\ell^\omega)_{[-r:]} u_\ell^{\omega X_1}
		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
		\quad .
	\end{align*}
	Remark that $\CutSuff{\ell,r}$ has all of its indeterminates
	multiplied by $\omega$, so that the induction hypothesis
	can be used.

@@

\begin{proofof}[polyrec-integer-strong:lem]
    Because $n P \in \CorrectPoly$,
    it is represented by a \kl{star-free $\Nat$-polyregular function} $f$ thanks to
    \cref{corrected-version:thm}. Let us now use \cref{n-sf-combinatorics:lem}
    to obtain a number $s \geq 1$ (that only depends on $P$) such
    that for all \kl{pumping patterns} $q \colon \Nat^p \to \Sigma^*$, $f
    \circ q(X_1 + s, \dots, X_p + s)$ is a \kl{natural binomial function} $F$
    over $\Nat^p$. We will use this $s$ with $p = k$, where
    $k = \card{\vec{X}}$ is the number of indeterminates of $P$,
    and $q(X_1, \dots, X_k) \defined \prod_{i = 1}^k a_i^{X_i}$,
    assuming the alphabet $\Sigma$ is $\set{a_1, \dots, a_k}$.
    In particular, $F(x_1, \dots, x_k) = f \circ q (x_1, \dots, x_k) = nP(x_1, \dots, x_k)$
    for all $\vec{x} \in (\Nat_{\geq s})^k$.
    Let us write the
    \kl{natural binomial function} $F$ as follows:
    \begin{equation*}
        F(X_1, \dots, X_k) =
        \sum_{\card{\alpha} \leq d} e_{\alpha} \prod_{i = 1}^k \binom{X -
        p_{\alpha,i}}{\alpha(i)} \quad
        \text{where } \forall 
        \card{\alpha} \leq d,
        e_{\alpha} \in \Nat
        \quad .
    \end{equation*}

    Because of the equality $\binom{X - p - 1}{k} + \binom{X - p - 1}{k -
    1} = \binom{X - p}{k}$, we conclude that we can find a $K \in \Nat$ and
    coefficients $f_{\alpha} \in \Nat$ such that $F(X_1, \dots, X_k) =
    \sum_{\card{\alpha} \leq d} f_{\alpha} \prod_{i = 1}^k \binom{X_i -
    K}{\alpha(i)}$, i.e., we can assume that $p_{\alpha,i} = K$ for all
    $\card{\alpha} \leq d$.
    Let us remark that $P$ is an \kl{integer-valued} polynomial of
    degree $d$, and therefore one can use
    \cref{integer-binomial-polynomial:lem} to rewrite it in the following form:
    \begin{equation*}
        P(X_1, \dots, X_k) = \sum_{\card{\alpha} \leq d}
        c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
        \quad \text{where } \forall \card{\alpha} \leq d, 
        c_{\alpha}
        \in \Rel \quad .
    \end{equation*}

    Because $F(x_1, \dots, x_k) = P(x_1, \dots, x_k)$ for all
    $\vec{x} \in (\Nat_{\geq s})^k$, one can write
    \begin{equation*}
        n \left(\sum_{\card{\alpha} \leq d} c_{\alpha} \prod_{i = 1}^k \pbinom{x_i - K}{\alpha(i)} \right)
        = 
        \sum_{\card{\alpha} \leq d} f_{\alpha} \prod_{i = 1}^k \binom{x_i - K}{\alpha(i)}
        \quad 
        \forall \vec{x} \in (\Nat_{\geq s})^k
    \end{equation*}

    Because this equality holds over $(\Nat_{\geq s})^k$, we conclude that the
    underlying polynomials are equal in $\Rel[\vec{X}]$, and therefore that:
    \begin{equation*}
        \sum_{\card{\alpha} \leq d} n c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
        = 
        \sum_{\card{\alpha} \leq d} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
    \end{equation*}

    Identifying the coefficients of the two polynomials (which is possible
    because the products of binomial coefficients translated by $K$ form a
    basis of $\Rel[\vec{X}]$), we conclude that for all $\card{\alpha} \leq d$,
    $n c_{\alpha} = f_{\alpha}$. In particular,
    for all $\card{\alpha} \leq d$, $c_{\alpha} \in \Nat$,
    hence $P$ is a \kl{natural binomial polynomial}.

    Remark that this reasoning remains true whenever we partially apply $P$ to
    some values in $\Nat$, since partial applications of $nP$ continue to be in
    $\CorrectPoly$ by definition, and remain \kl{integer-valued}. We have
    proven that $P$ is a \kl{strongly natural binomial polynomial}.
\end{proofof}

\section{Proofs of section  \ref{beyond-polynomials:sec}}

\begin{proofof}[decompose-polynomial:lem]
    Let us first remark that one only needs to consider the case where $f$ is a
    function in $\NPoly$, because every function $f \in \ZPoly$ can be decomposed
    as $f = f_+ - f_-$ where $f_+$ and $f_-$ 
    are \kl{$\Nat$-polyregular functions}, and because \kl{integer binomial
    polynomials} are closed under subtraction.
    Now, let us prove this result by induction on the number of letters in the alphabet.
    When the alphabet is empty, the result is trivial because $f$ can only be applied to $\varepsilon$,
    hence can be represented using a constant polynomial.

    Assume that the result holds for all functions with an alphabet of size at
    most $k-1$. We first apply \cref{n-poly-combinatorics:lem} to obtain a number
    $\omega$ such that for all \kl{pumping patterns} $q \colon \Nat^\ell \to
    \Sigma^*$, there exists a (computable) \kl{natural binomial function} $F_q$
    such that $f \circ q (\omega X_1, \dots, \omega X_\ell) = F_q$ over
    $(\Nat_{\geq 1})^\ell$.
    Given a tuple $\vec{r} \in \set{0, \dots, \omega - 1}^k$, and a 
    set $S \subseteq \set{1, \dots, k}$,
    let us define
    the \kl{pumping pattern} $q_{(S,\vec{r})} \colon \Nat^{\card{S}} \to \Sigma^*$ as
    $q_{(S, \vec{r})}(\seqof{X_i}{i \in S}) \defined \prod_{i = 1}^k a_i^{r_i} \cdot \prod_{i \in S} a_i^{X_i}$.
    Let us write $F_{(S,\vec{r})} \colon \Nat^{\card{S}} \to \Nat$ 
    for the associated
    \kl{natural binomial function} satisfying:
    \begin{equation*}
        f \left(
            \prod_{i = 1}^k a_i^{r_i}
            \cdot
            \prod_{i \in S} a_i^{\omega x_i}
        \right)
        = 
        F_{(S,\vec{r})}(\seqof{x_i}{i \in S})
        \quad 
        \forall \vec{x} \in (\Nat_{\geq 1})^{S}
        \quad .
    \end{equation*}

    Such a function $F_{(S,\vec{r})}$ coincides with a \kl{natural binomial
    polynomial} $P_{(S,\vec{r})} \in \Rat[\seqof{X_i}{i \in S}]$ on large
    enough inputs. Let us write $N_0 \geq 1$ be a number such that
    $F_{(S,\vec{r})}(\seqof{x_i}{i \in S}) = P_{(S,\vec{r})}(\seqof{x_i}{i \in
    S})$ for all $(S, \vec{r}) \in \ModuloTypes[\omega]^k$, for all $\vec{x}
    \in (\Nat_{\geq N_0})^{S}$. This $N_0$ is computable because one can
    collect upper bounds from the syntax of the \kl{natural binomial functions}
    $F_{(S,\vec{r})}$.
    We conclude that:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^k a_i^{x_i}
        \right)
        = 
        P_{(S, \vec{r})}(\seqof{\floor{x_i / \omega}}{i \in S})
        \quad 
        \text{ where } (S, \vec{r}) = \moduloType[\omega](\vec{x})
        \text{ and }
        \vec{x} \in (\Nat_{\geq N_0 \omega})^k
        \quad .
    \end{equation*}

    This provides a representation for $f$ on large enough inputs, and we can
    apply the induction hypothesis on the functions $f_{a^n} \colon (\Sigma
    \setminus \set{a})^* \to \Nat$ in $\NPoly$ defined by $f_{a^n} (w) \defined
    f(a^n \cdot w)$, for every $a \in \Sigma$ and $n \leq N_0\omega$. Remark
    that \kl{integer binomial polynomials} are closed multiplying the input by
    a constant, hence that we can always move from a decomposition using
    $\omega$ to a decomposition using a multiple of $\omega$.
    Let us then consider as our new $\omega'$ the least common multiple of
    $\omega$, $N_0$, and all the $\omega_{a^n}$'s obtained for each $f_{a^n}$.

    We can now, given an entry $\vec{x} \in \Nat^k$ with
    $\moduloType[\omega'](x) = (S, \vec{r})$ recognize if a some letter
    $a$ appears less than $N_0 \omega$ times in the input, and apply the
    corresponding polynomials from the induction hypothesis if it is the case.
    Whenever all letters appear at least $N_0 \omega$ times, we can apply the
    polynomial $P_{(S,\vec{r})}$.
\end{proofof}

### 2024-11-29 parts/prelim.tex
big corrections

\begin{lemma}
    \label{decidable-commutative-poly:lemma}
    \label{decidable-commutative-rat:lemma}
    Let $f \in \ZRat$. One can decide if 
    $f$
    is \kl{commutative}.
\end{lemma}
\begin{proof}
    Remark that the group of permutations of $\set{1, \dots, n}$ is generated by
    the cycle $c \defined (n,1, \dots, n-1)$ and the transposition $t \defined (1, 2)$.
    As a consequence, a function $f$ is commutative if and only if
    $f \circ c = f = f \circ t$.
    When $f$ is a \kl{rational series},
    $f \circ c$ and $f \circ t$ are both \kl{rational series} that can be
    effectively computed from $f$ \textbf{TODO cite}, and since equivalence
    of rational series is decidable 
    \cite[Corollary 3.6]{BERE10},
    we have obtained a decision procedure.
\end{proof}

>>>>

\begin{lemma}
    \label{decidable-commutative-poly:lemma}
    \label{decidable-commutative-rat:lemma}
    Let $f \in \ZRat$. One can decide if 
    $f$
    is \kl{commutative}.
\end{lemma}
\begin{proof}
    Remark that the group of permutations of $\set{1, \dots, n}$ is generated by
    the cycle $c \defined (n,1, \dots, n-1)$ and the transposition $t \defined (1, 2)$.
    As a consequence, a function $f$ is commutative if and only if
    $f \circ c = f = f \circ t$.
    When $f$ is a \kl{rational series},
    $f \circ c$ and $f \circ t$ are both \kl{rational series} that can be
    effectively computed from $f$,\footnote{
        This can be done by guessing the second (resp. last) letter of the input word, 
        remembering the first letter in a state, and 
        then running the original automaton for $f$ on the modified input, checking at the second position 
        (resp. the end of the word)
        if the guess was correct.
    } and since equivalence
    of rational series is decidable 
    \cite[Corollary 3.6]{BERE10},
    we have obtained a decision procedure.
\end{proof}
